Networks Matter in International Relations:\\A Latent Network Approach to Dyadic Analysis

### 2017-10-24 paper/4_app.tex
small stuff.

\section{Re-estimation with AME}

@@

We obtained the data for each of these studies from their replication archives and replicated the main results of each of the articles.\footnote{Without exception this was straightforward to accomplish, thanks to an increasing norm in the social sciences of open data sharing.} We also examine each of the models using the AME framework described above.  Our goal is to ascertain whether the ignored interdependencies --- the non-iid structure of the relational data --- would result in different model estimates when they were addressed in an AME framework, and more importantly to see if there were substantive opportunities that were presented with the dynamic factor approach.  
Finally, in each study we assess whether there is any new substantive finding that emerges or indeed if any disappear once the interdependencies in the data are modeled.

@@

% SM: maybe too on the nose (Whether putative control variables, for example, are precisely estimated is less important.) 
Given current practices, most scholars have a single variable in a complicated empirical model that they look at for evaluating the validity of their empirically estimated models. Generally, interpretations focus on a small set of independent variables.  For example, in the study by Reiter and Stam (2003) the goal is to determine whether in mixed dyads --- consisting of a democratic country and country ruled by a personalist dictatorship --- it is the democratic country or the personalist country that is associated with a greater likelihood of the dyad being in a dispute.  In a regression including fourteen covariates, they focus on whether the variables that capture whether the initiator is a democratic country are statistically significant.\footnote{In particular, the variables \texttt{Pers/Democ} and \texttt{Democ/Pers} in Model 3, in Table 1, page 335.} In our re-examination of Reiter and Stam, we focus on whether the model replicates when re-estimated within the AME framework. Each of the other four studies also has a crucial finding that we hone in on to further draw into focus the potential differences brought about by utilizing an AME estimation procedure.  In Table~\ref{tab:modelFindingSumm}, we present the overall results; the term \textit{does not replicate} indicates only that the sign and/or significance of the putatively crucial finding in the original study is not found to hold in the AME estimation.

@@

\noindent The result of this procedure is a set of sociomatrices $\bm \hat Y$, in which each entry $\hat y_{ij,t}$ is a predicted value obtained from using a subset of the data that does not include $y_{ij,t}$.\footnote{For more details on this type of cross-validation strategy see \citet{minhas:etal:2016:arxiv}.} We summarize the performance of the various models in Table~\ref{tab:modelPerfSumm} below. For the binary models we just provide the area under the Receiver Operator Characteristic (ROC) and Precision Recall (PR) curves. Only one of the studies here had a continuous dependent variable and for this we provide the root mean squared error (RMSE) and root median squared error (RMDSE).\footnote{More details on the performance of each of these models can be found in the Appendix.} For each of the replications, we find that the AME approach substantially outperforms the original models in terms of out of sample predictive performance.

@@

In 2004, Andrew Rose published a study in the \textit{American Economic Review} \nocite{rose:2004} that proved to be quite controversial in terms of macroeconomic trade theory and in terms of trade policy in a variety of nations. It also provoked a number of responses in the international political economy literature \cite{tomz:etal:2007,ward:etal:2013}.  Rose's basic argument is that despite longstanding arguments, made by trade theorists and the World Trade Organization, that WTO membership fosters greater cooperation and thereby more trade among its members, the empirics do not bear out such claims. He uses a standard gravity model with dyadic data on bilateral merchandise trade (not services) for $175$ countries over a period of five decades. Estimating this model using OLS within many differing contexts, his conclusion was that: ``An extensive search reveals little evidence that countries joining or belonging to the GATT/WTO have different trade patters from outsiders \ldots (2004, page 98, abstract).''  The data for this study have been widely used in replications by many searching for the missing effects of the WTO --- as well as preferential trade agreements, bilateral investment theories, and other aspects of modern trade theory.

When we compare the results of Rose's original OLS model to the AME model accounting for network dependencies, the results are generally similar. The main result of the model --- the null effect of membership in the WTO, as represented by the ``One-In" and ``Both-In" variables --- persists in the AME model. However, while in the original model there was a clear positive relationship between Real GDP and Trade, most of this effect vanishes in the AME model. The random effects shown in Figure~\ref{fig:roser} reveal the cause of much of this divergence. Here, the states with the most positive random effects are also states with high GDP, though not necessarily high GDP/capita.\footnote{Note: Qatar exhibits strongly negative random effects.} Thus, the effect of GDP in the original model was, in part, an artifact of first-order dependencies. Most of the other results of the model are constant across each model, though some geographic features, such as islands and landlocked states, have a more clear effect on trade once we account for these network dependencies. When we account for network interdependencies, we observe a markedly lower Root Mean Squared Error out-of-sample --- $3.23$ for the OLS model and $1.77$ for the AME model. While the AME model replicate the main result, there are substantial and substantive improvements gleaned from the AME results.

@@

The re-estimation of \citet{weeks:2012} likely has the sharpest divergence between the General Linear Model results and those of the AME Model. In Weeks's initial models,  she finds that machines are less prone to initiate conflict than the reference category, whereas Juntas, Bosses and Strong-men are more conflict-prone, in line with her theoretical expectations. When we look at the posterior distribution of these coefficients in the AME results, we find that none of these values are distinguishable from zero. Similarly, we find less pronounced effects for military capabilities. One explanation for this divergence is the AME model's ability to account for third order effects. Inspection of the multiplicative effects in Figure~\ref{fig:weekscirc} reveals a number of clusters of states which exhibit structural equivalence --- in the top right corner we see states like the US, the UK, and Israel who often send conflict to similar targets. We also observe a cluster of ``rogue states,'' who are receivers of conflict, in the top right (Iran, Iraq, Syria, Libya, North Korea). In the bottom corner we see a cluster of authoritarian senders including Iraq, Russia, Syria, North Korea and China. In general, these clusters have similar governmental types (Iraq, Syria, Libya, and North Korea all fell under the ``boss" category), and similar conflict behavior. In the GLM, which ignores these third order dependencies, many of these results might have been attributed to regime type. The AME model, on the other hand, shows that this can be specified in terms of the interdependencies captured by the multiplicative effects. In terms of out-of-sample performance, shown in Figure~\ref{fig:weeksauc}, the AME model performs markedly better than the GLM out of sample, lending credence to the possibility that these third-order dependencies cause spurious effects for regime type.  In terms of performance, the original Weeks' model does slightly better than a coin-toss overall, and fails to predict more than a couple of events (incorrectly) in almost one million observations.

@@

We re-estimated model 6 from Table 6 (2017, 34). The results are presented in the Appendix. The results obtained with AME stand in stark contrast to those found with a logistic regression (with dyad clustered, robust standard errors).  Most importantly, the primary variable from the Gibler study, parity of the members of a dyad at the year in which they entered the international system, is shown to be unimportant in the AME results.  Not only is the value of this parameter small, but it has a very large relative standard error --- over a magnitude larger than the parameter itself ($z= 0.038$). In addition, the variable indicating whether both members of the dyad were coded as democracies (joint democracy) follows the same pattern: important and strong in the logistic results, but this disappears once interdependencies are modeled.  As might be expected, the strong geographic clustering in the original study is about one-quarter as strong in the AME estimations. Similarly, rivalry coefficients are about one-third the size in the AME formulation, but a great deal more precisely measured ($z=18.116$).

We utilized the original and the AME results from Gibler's model 6 in Table 6 to calculate the expected values for one scenario. We focused on the variables measuring rivalry.  We used mean or modal values for all independent variables, except we changed the rivalry variable to indicate that there was a rivalry when the actual data suggest there is none.  The expected values of this scenario are essentially a first difference plot comparing results with the model when estimated in two different ways: Gibler's GLM estimation and our AME approach.  As this Figure~\ref{fig:gibmargeff} illustrates, the AME results differ notable. First, the expected value of the dependent variable --- the probability of the onset of a militarized interstate dispute, is considerably lower when taking interdependencies into account with the AME model.  These are rare events, so the probabilities are low, but the difference is a factor of almost 2. Thus, you get quite substantially different expected values from these two models.

@@

First and foremost, many findings which emerge from models that do not take interdependencies into account lose their statistical significance when network effects are estimated via AME.\footnote{This should not be news --- since the finding has long been in the theoretical literature, but given the state of current literature in international relations is still pertinent.} Not only are coefficients biased in the GLM approaches to the analysis of dyadic data, but they are often imprecisely measured, with inflated standard errors.  This means that significant testing (for better or worse) is compromised when network effects are ignored.

@@

Third, it is evident that the actual results --- not the estimated coefficents and their covariances --- which are generated by the models differ greatly in expectations.  This implies that policy experimentations with the models, as well as scenario-based simulations and forecasting of GLM models are likely to often give misleading results compared to the AME approach.

>>>>

\section{\textbf{Re-estimation with AME}|

@@

We obtained the data for each of these studies from their replication archives and replicated the main results of each of the articles.\footnote{Without exception this was straightforward to accomplish, thanks to an increasing norm in the social sciences of open data sharing.} We also examine each of the models using the AME framework described above.  Our goal is to ascertain whether the ignored interdependencies---the non-iid structure of the relational data---would result in different model estimates when they were addressed in an AME framework, and more importantly to see if there were substantive opportunities that were presented with the dynamic factor approach.  
Finally, in each study we assess whether there is any new substantive finding that emerges or indeed if any disappear once the interdependencies in the data are modeled.

@@

% SM: maybe too on the nose (Whether putative control variables, for example, are precisely estimated is less important.) 
Given current practices, most scholars have a single variable in a complicated empirical model that they look at for evaluating the validity of their empirically estimated models. Generally, interpretations focus on a small set of independent variables.  For example, in the study by Reiter and Stam (2003) the goal is to determine whether in mixed dyads---consisting of a democratic country and country ruled by a personalist dictatorship---it is the democratic country or the personalist country that is associated with a greater likelihood of the dyad being in a dispute.  In a regression including fourteen covariates, they focus on whether the variables that capture whether the initiator is a democratic country are statistically significant.\footnote{In particular, the variables \texttt{Pers/Democ} and \texttt{Democ/Pers} in Model 3, in Table 1, page 335.} In our re-examination of Reiter and Stam, we focus on whether the model replicates when re-estimated within the AME framework. Each of the other four studies also has a crucial finding that we hone in on to further draw into focus the potential differences brought about by utilizing an AME estimation procedure.  In Table~\ref{tab:modelFindingSumm}, we present the overall results; the term \textit{does not replicate} indicates only that the sign and/or significance of the putatively crucial finding in the original study is not found to hold in the AME estimation.

@@

\noindent The result of this procedure is a set of sociomatrices $\bm \hat Y$, in which each entry $\hat y_{ij,t}$ is a predicted value obtained from using a subset of the data that does not include $y_{ij,t}$.\footnote{For more details on this type of cross-validation strategy see \citet{minhas:etal:2016:arxiv}.} We summarize the performance of the various models in Table~\ref{tab:modelPerfSumm} below. For the binary models we just provide the area under the Receiver Operator Characteristic (ROC) and Precision Recall (PR) curves. Only one of the studies here had a continuous dependent variable and for this we provide the root mean squared error (RMSE) and root median squared error (RMDSE).\footnote{More details on the performance of each of these models can be found in the Appendix.} For each of the replications, we find that the AME approach substantially outperforms the original models in terms of out-of-sample predictive performance.

@@

In 2004, Andrew Rose published a study in the \textit{American Economic Review} \nocite{rose:2004} that proved to be quite controversial in terms of macroeconomic trade theory and in terms of trade policy in a variety of nations. It also provoked a number of responses in the international political economy literature \cite{tomz:etal:2007,ward:etal:2013}.  Rose's basic argument is that despite longstanding arguments, made by trade theorists and the World Trade Organization, that WTO membership fosters greater cooperation and thereby more trade among its members, the empirics do not bear out such claims. He uses a standard gravity model with dyadic data on bilateral merchandise trade (not services) for $175$ countries over a period of five decades. Estimating this model using OLS within many differing contexts, his conclusion was that: ``An extensive search reveals little evidence that countries joining or belonging to the GATT/WTO have different trade patters from outsiders \ldots (2004, page 98, abstract).''  The data for this study have been widely used in replications by many searching for the missing effects of the WTO---as well as preferential trade agreements, bilateral investment theories, and other aspects of modern trade theory.

When we compare the results of Rose's original OLS model to the AME model accounting for network dependencies, the results are generally similar. The main result of the model---the null effect of membership in the WTO, as represented by the ``One-In" and ``Both-In" variables---persists in the AME model. However, while in the original model there was a clear positive relationship between Real GDP and Trade, most of this effect vanishes in the AME model. The random effects shown in Figure~\ref{fig:roser} reveal the cause of much of this divergence. Here, the states with the most positive random effects are also states with high GDP, though not necessarily high GDP/capita.\footnote{Note: Qatar exhibits strongly negative random effects.} Thus, the effect of GDP in the original model was, in part, an artifact of first-order dependencies. Most of the other results of the model are constant across each model, though some geographic features, such as islands and landlocked states, have a more clear effect on trade once we account for these network dependencies. When we account for network interdependencies, we observe a markedly lower Root Mean Squared Error out-of-sample --- $3.23$ for the OLS model and $1.77$ for the AME model. While the AME model replicate the main result, there are substantial and substantive improvements gleaned from the AME results.

@@

The re-estimation of \citet{weeks:2012} likely has the sharpest divergence between the General Linear Model results and those of the AME Model. In Weeks's initial models,  she finds that machines are less prone to initiate conflict than the reference category, whereas Juntas, Bosses and Strong-men are more conflict-prone, in line with her theoretical expectations. When we look at the posterior distribution of these coefficients in the AME results, we find that none of these values are distinguishable from zero. Similarly, we find less pronounced effects for military capabilities. One explanation for this divergence is the AME model's ability to account for third order effects. Inspection of the multiplicative effects in Figure~\ref{fig:weekscirc} reveals a number of clusters of states which exhibit structural equivalence---in the top right corner we see states like the US, the UK, and Israel who often send conflict to similar targets. We also observe a cluster of ``rogue states,'' who are receivers of conflict, in the top right (Iran, Iraq, Syria, Libya, North Korea). In the bottom corner we see a cluster of authoritarian senders including Iraq, Russia, Syria, North Korea and China. In general, these clusters have similar governmental types (Iraq, Syria, Libya, and North Korea all fell under the ``boss" category), and similar conflict behavior. In the GLM, which ignores these third order dependencies, many of these results might have been attributed to regime type. The AME model, on the other hand, shows that this can be specified in terms of the interdependencies captured by the multiplicative effects. In terms of out-of-sample performance, shown in Figure~\ref{fig:weeksauc}, the AME model performs markedly better than the GLM out-of-sample, lending credence to the possibility that these third-order dependencies cause spurious effects for regime type.  In terms of performance, the original Weeks' model does slightly better than a coin-toss overall, and fails to predict more than a couple of events (incorrectly) in almost one million observations.

@@

We re-estimated model 6 from Table 6 (2017, 34). The results are presented in the Appendix. The results obtained with AME stand in stark contrast to those found with a logistic regression (with dyad clustered, robust standard errors).  Most importantly, the primary variable from the Gibler study, parity of the members of a dyad at the year in which they entered the international system, is shown to be unimportant in the AME results.  Not only is the value of this parameter small, but it has a very large relative standard error---over a magnitude larger than the parameter itself ($z= 0.038$). In addition, the variable indicating whether both members of the dyad were coded as democracies (joint democracy) follows the same pattern: important and strong in the logistic results, but this disappears once interdependencies are modeled.  As might be expected, the strong geographic clustering in the original study is about one-quarter as strong in the AME estimations. Similarly, rivalry coefficients are about one-third the size in the AME formulation, but a great deal more precisely measured ($z=18.116$).

We utilized the original and the AME results from Gibler's model 6 in Table 6 to calculate the expected values for one scenario. We focused on the variables measuring rivalry.  We used mean or modal values for all independent variables, except we changed the rivalry variable to indicate that there was a rivalry when the actual data suggest there is none.  The expected values of this scenario are essentially a first difference plot comparing results with the model when estimated in two different ways: Gibler's GLM estimation and our AME approach.  As this Figure~\ref{fig:gibmargeff} illustrates, the AME results differ notable. First, the expected value of the dependent variable---the probability of the onset of a militarized interstate dispute, is considerably lower when taking interdependencies into account with the AME model.  These are rare events, so the probabilities are low, but the difference is a factor of almost 2. Thus, you get quite substantially different expected values from these two models.

@@

First and foremost, many findings which emerge from models that do not take interdependencies into account lose their statistical significance when network effects are estimated via AME.\footnote{This should not be news---since the finding has long been in the theoretical literature, but given the state of current literature in international relations is still pertinent.} Not only are coefficients biased in the GLM approaches to the analysis of dyadic data, but they are often imprecisely measured, with inflated standard errors.  This means that significant testing (for better or worse) is compromised when network effects are ignored.

@@

Third, it is evident that the actual results---not the estimated coefficents and their covariances---which are generated by the models differ greatly in expectations.  This implies that policy experimentations with the models, as well as scenario-based simulations and forecasting of GLM models are likely to often give misleading results compared to the AME approach.

### 2017-10-24 paper/6_appendix.tex
small stuff.

Additionally, for each replication we provide a more detailed visualization illustrating the results of our out of sample performance analysis. For Weeks (2012) we already did so in the main body of the text so we eschew from doing so again here.

@@

If directed, AMEN further calls for:

>>>>

Additionally, for each replication we provide a more detailed visualization illustrating the results of our out-of-sample performance analysis. For Weeks (2012) we already did so in the main body of the text so we eschew from doing so again here.

@@

If directed, AMEN further requires:
